---
title: "Phase transitions in software repositories"
author: "JJ Merelo"
date: "20 de febrero de 2017"
output:
  pdf_document: default
  html_document: default
bibliography: biblio.bib
subtitle: Examining ggplot2
abstract: This document provides an overview of phase transitions in sofware repositories
  examining ggplot2, an open source R graphics and visualization library in active development. By
  examining how the file co-modification graph evolves, it shows how phase transitions take the repository
  to a critical state through the self-organization of the software developers, since there is no central authority that organizes it. 
---

```{r setup, include=FALSE}
library(ggplot2)
load(file="../git-mining/data/edges-ggplot2-data.Rda")
```

# Introduction

The fact that software projects as registered in software repositories constitute complex systems has been known for a long time @lopez2008applying, @lopez2004applying  @Merelo2016repomining. Since the beginning of the century several authors have examined software repositories and looked at size of changes, file coauthorship graph, and other types of graphs, concluding that they are a scale free system as reflected by many different measures such as the degree of coauthorship in author graphs or the size of changes.

The main instruments for taking these measures have been co-authorship graphs for modules or even files. However, The main problem with this is that they tend to create densely connected graphs since there are files such as the documentation or build tools or modules such as the test folder where most authors have contributed. To make these file coauthorship graphs *interesting*, some filters have to be applied eliminating those files, for instance, or considering only certain type of files. 

File graphs do have the same problem. A repository is a bipartite graph of files and authors, and *flattening* it to any of them reflects the same problem. Some authors have worked in many files, which means that there are files densely connected forming cliques that have little, or nothing, to do with the complex nature of the system.

We are interested in looking for a graph representation whose measures and evolution actually reflect the fact that through a self-organized phase transition they have reached a critical state @doi101093comnetcnt022. That is why we will be looking at a type of graph, the graph of co-ocurrence in commits, which we have examined before in @vue2017, looking for increments in measures that reflect this phase transitions to a critical state. 

## Methodology

Data has been extracted from GitHub repositories and examining the log. A graph has been generated with vertices in every file and edges that connect two files if they appear in the same commit at least once. This has been done with a Perl script which can be downloaded [from GitHub](http://github.com/JJ/git-mining). Every edge is labelled with the sequence index of the commit it appears. An R script examines this file and then computes network measures for the accumulated graph at that particular point in time. The size of the giant connected component, the average degree, betweenness and transitivity (clustering coefficient) are computed. This will be shown below. 

The repository has a total of around 4000 commits. Not in all commits new edges are added. The evolution of these measures will be shown below.

## Finding phase transitions 

We are looking for phase transitions in these systems. At one or possible several points in time, the graph might *collapse*, giving way to the long-distance correlations that have been observed. One way of looking at this is plotting the evolution of the average degree.

```{r degree, echo=FALSE}
ggplot(data=measures,aes(x=commit,y=degree))+ geom_line()
```

```{r betweenness, echo=FALSE}
ggplot(data=measures,aes(x=commit,y=betweenness))+ geom_line()
```

```{r closeness, echo=FALSE}
ggplot(data=measures,aes(x=commit,y=closeness))+ geom_line()
```

```{r connected.rate, echo=FALSE}
ggplot(data=measures,aes(x=commit,y=connected.rate))+ geom_line()
```

```{r diameter, echo=FALSE}
ggplot(data=measures,aes(x=commit,y=diameter))+ geom_line()
```

```{r distance, echo=FALSE}
ggplot(data=measures,aes(x=commit,y=distance))+ geom_line()
```

```{r clustering, echo=FALSE}
ggplot(data=measures,aes(x=commit,y=transitivity))+ geom_line()
```


# Conclusions



# References
